Bacterial Microbiome and Obesity Meta-analysis Supplemental

### 2016-02-12 code/microbiomeAndobesitySupplemental.Rmd
manuscript update 2

---
title: "Bacterial Microbiome and Obesity Meta-analysis Statistics Pipeline"
author: "Marc Sze"
date: "January 29, 2016"
output: 
  html_document:
    keep_md: yes

---
<br>

#### **Preamble**  
The following workflow was used to generate all the statistical analysis of our meta-analysis.  It is organized by part of which there are three.  Code chunks will be displayed as such below:

```{r,  eval=FALSE}
print("Hello World!")
x = 1 + 1
print(x)
```
Data tables are provided in this repository for your use if you do not want to have to repeat the somewhat laborious sequence analysis pipeline from start to finish.  BMI data if publicly available has been kept in this repository for your ease of reference.  However, there are a few data sets in which the BMI data is not public and as such you will need to request this information from the respective sources if you would like to use them.  The two data sets this mostly concerns are the Wu, et al study and the HMP study.
<br><br>

#### **Part 1: Seperate Analysis of each Study**  
The statistical analysis will follow a general order of data loading, alpha diversity and BMI analysis, PERMANOVA of Bray-Curtis distance matrix on obese versus non-obese and BMI classifications (where available), relative risk, and then classification using AUCRF.
<br><br>

##### **Baxter et al. 2016:**  
```{r,eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

# Reading in the necessary Data
setwd("C:/Users/marc/Desktop/obesity2/data1")
demographics <- read.csv("demographics.v2.csv")
microbiome <- read.csv("data1.subsample.otus.csv")
phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
taxonomy <- read.csv("data1.taxonomy.csv")

# Minor modifications and Rownames adustments of data tables
rownames(demographics) <- demographics[,1]
rownames(microbiome) <- microbiome[,2]
demographics <- demographics[,-1]
microbiome <- microbiome[,-2]
test.samples <- rownames(demographics)
microbiome <- microbiome[,-c(1:2)]

# Create Table for phyla information
phylogenetic.info <- phylogenetic.info[-c(1:12),]
phyla <- which(phylogenetic.info$taxlevel == 2)
phyla.table <- phylogenetic.info[phyla, ]
phyla.table <- phyla.table[,-c(1:2,4:5)]
phyla.names <- as.character(phyla.table[,1])
phyla.table <- phyla.table[,-1]
phyla.table <- as.data.frame(t(phyla.table))
colnames(phyla.table) <- phyla.names
rownames(phyla.table) <- rownames(demographics)

#Add phyla together that are not very abundant and delete them from the table
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Generate Alpha Diversity Measures and alpha diversity table
library(vegan)
H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Create age quartiles

demographics$age.quartile <- with(demographics, cut(Age, breaks=quantile(Age, probs=seq(0,1, by=0.25)), include.lowest=TRUE))

#Create Obese Yes/No groups
demographics$obese[demographics$BMI.classification=="Normal" | demographics$BMI.classification=="Overweight"] <- "No"
demographics$obese[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Yes"

#Create Obese.num groups
demographics$obese.num[demographics$obese=="No"] <- 0
demographics$obese.num[demographics$obese=="Yes"] <- 1
obese <- factor(demographics$obese.num)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

#Create a column with obese and extreme obese as single entity
demographics$BMIclass2[demographics$BMI.classification=="Normal"] <- "Normal"
demographics$BMIclass2[demographics$BMI.classification=="Overweight"] <- "Overweight"                     
demographics$BMIclass2[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Obese"

##Test BMI versus alpha diversity and phyla

bmi <- demographics$BMI
bmi.class <- factor(demographics$BMI.classification)
bmi.class2 <- factor(demographics$BMIclass2)

anova(lm(H ~ obese)) #P-value=0.01643
anova(lm(H ~ bmi.class)) #P-value=0.06
anova(lm(H ~ bmi.class2)) #P-value=0.04471
a2 <- aov(H ~ bmi.class2)
TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
# Obese-Norm (p-value=0.201), Over-Norm (p-value=0.769), Over-Obese (p-value=0.036)

anova(lm(S ~ obese)) #P-value=0.01571
anova(lm(S ~ bmi.class)) #P-value=0.06
anova(lm(S ~ bmi.class2)) #P-value=0.03
a2 <- aov(S ~ bmi.class2)
TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
# Obese-Norm (p-value=0.025), Over-Norm (p-value=0.546), Over-Obese (p-value=0.178)

anova(lm(J ~ obese)) #P-value=0.03241
anova(lm(J ~ bmi.class)) #P-value=0.0272
a2 <- aov(J ~ bmi.class)
TukeyHSD(x=a2, 'bmi.class', conf.level=0.95)
# Obese-Norm (p-value=0.02), Over-Norm (p-value=0.047), Over-Obese (p-value=0.06)
anova(lm(J ~ bmi.class2)) #P-value=0.02
a2 <- aov(J ~ bmi.class2)
TukeyHSD(x=a2, 'bmi.class2', conf.level=0.95)
# Obese-Norm (p-value=0.630), Over-Norm (p-value=0.159), Over-Obese (p-value=0.019)

##linear correlations
H.bmi.test <- lm(H ~ bmi)
summary(H.bmi.test)  
#R2 = 0.01246, P-value=0.0774

S.bmi.test <- lm(S ~ bmi)
summary(S.bmi.test)
#R2 = 0.03222, P-value=0.0105

J.bmi.test <- lm(J ~ bmi)
summary(J.bmi.test)
#R2 = -0.0003052, P-value=0.332

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.3175
anova(lm(bacter ~ bmi.class)) #P-value=0.6457
anova(lm(bacter ~ bmi.class2)) #P-value=0.5993

anova(lm(firm ~ obese)) #P-value=0.6817
anova(lm(firm ~ bmi.class)) #P-value=0.9293
anova(lm(firm ~ bmi.class2)) #P-value=0.8936

anova(lm(BFratio ~ obese)) #P-value=0.6305
anova(lm(BFratio ~ bmi.class)) #P-value=0.8334
anova(lm(BFratio ~ bmi.class2)) #P-value=0.881

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

table(demographics[,23])
###Normal=54, Overweight=71, Obese=45, Extreme Obesity=2
###Need to omit extreme obesity
#Removing rows with extreme obesity
test <- which(demographics$BMI.classification == "Extreme Obesity")
BMI.class.demo <- demographics[-test,]
BMI.microbiome <- microbiome[-test,]
bmi.class.group <- factor(BMI.class.demo$BMI.classification)

set.seed(3)
adonis(BMI.microbiome ~ bmi.class.group, permutations=1000)
#Not significant PERMANOVA = 0.2278, pseudo-F = 1.1323

set.seed(3)
adonis(microbiome ~ bmi.class, permutations=1000)
#Not significant PERMANOVA = 0.1848, pseudo-F = 1.1368

set.seed(3)
adonis(microbiome ~ bmi.class2, permutations=1000)
#Not significant PERMANOVA = 0.1359, pseudo-F = 1.1937

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe

#Shannon diversity
median(H) # 3.67469
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 198
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.692766
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Run the RR for Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:86), 4])
table(test3[c(87:172), 4])
#Group1 (Higher than median), obese = 22 and non-obese = 64
#Group2 (Lower than median), obese = 25 and non-obese = 61

group1 <- c(22, 64)
group2 <- c(25, 61)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.14
## CI = 0.70, 1.85
## p-value = 0.608

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.3621609
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:86), 2]) #higher group
table(test4[c(87:172), 2]) #lower group
#Group1 (Higher than median), obese = 23 and non-obese = 63
#Group2 (Lower than median), obese = 24 and non-obese = 62

group1 <- c(23, 63)
group2 <- c(24, 62)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.04
## CI = 0.64, 1.70
## p-value = 0.864

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#generate test set
  # get rid of values that only have 0 and something else
testset <- Filter(function(x)(length(unique(x))>2), microbiome)
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)

#Need to add phyla and alpha diversity measures to the dataset

testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
testset <- cbind(obese, testset)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 14 Measures, AUCopt = 0.7553
plot(fit)

```
<br><br>

##### **Ross et al. 2015:**  
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/hispanic/")

hispanic.microb <- read.table("hispanic.subsample.shared", header=T)

metadata <- read.csv("s40168-015-0072-y-s1.csv")
sample.match <- read.csv("Hispanic_dataset.csv")

#Create a microbiome data table with sample names from metadata
test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
test <- test[order(test[, 1]), ]

test2 <- cbind(test, hispanic.microb)
rownames(test2) <- test2[, 2]
his.microb.edit <- test2[, -c(1:5)]
edit.metadata <- metadata[, -c(2:5)]
rownames(edit.metadata) <- edit.metadata[, 1]
edit.metadata <- edit.metadata[, -1]

#Create a metadata file in the order of the microbiome data
order1 <- rownames(his.microb.edit)
edit.metadata2 <- edit.metadata[order1, ]

#Get alpha diversity of the samples
library(vegan)

H <- diversity(his.microb.edit)
S <- specnumber(his.microb.edit)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(his.microb.edit)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- his.microb.edit
colnames(phyla.table) <- phyla.names
  #add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
  #combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Deinococcus-Thermus", "Elusimicrobia", "Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(3:4, 6:7, 9:10)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- as.data.frame(phyla.table.rel.abund[, -8])

#Create BMI groups
edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"

#Create BMI class 2 groups
edit.metadata2$BMI.class2[edit.metadata2$BMI<=24] <- "Normal"
edit.metadata2$BMI.class2[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
edit.metadata2$BMI.class2[edit.metadata2$BMI>=30] <- "Obese"

#Get paitent demographics to be tested

bmi <- edit.metadata2$BMI
obese <- factor(edit.metadata2$obese)
bmi.class <- factor(edit.metadata2$BMI.class)
bmi.class2 <- factor(edit.metadata2$BMI.class2)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.1657
anova(lm(H ~ bmi.class)) #P-value=0.5448
anova(lm(H ~ bmi.class2)) #P-value=0.3615

anova(lm(S ~ obese)) #P-value=0.1841
anova(lm(S ~ bmi.class)) #P-value=0.61
anova(lm(S ~ bmi.class2)) #P-value=0.4

anova(lm(J ~ obese)) #P-value=0.2538
anova(lm(J ~ bmi.class)) #P-value=0.6431
anova(lm(J ~ bmi.class2)) #P-value=0.492

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.343, R2=0.01476
summary(lm(S ~ bmi)) #P-value=0.345, R2=0.01466
summary(lm(J ~ bmi)) #P-value=0.472, R2=0.008502

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.2092
anova(lm(bacter ~ bmi.class)) #P-value=0.5436
anova(lm(bacter ~ bmi.class2)) #P-value=0.4374

anova(lm(firm ~ obese)) #P-value=0.2569
anova(lm(firm ~ bmi.class)) #P-value=0.1741
anova(lm(firm ~ bmi.class2)) #P-value=0.321

anova(lm(BFratio ~ obese)) #P-value=0.2508
anova(lm(BFratio ~ bmi.class)) #P-value=0.4966
anova(lm(BFratio ~ bmi.class2)) #P-value=0.506

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(his.microb.edit ~ obese, permutations=1000) 
  #PERMANOVA=0.7153, pseudo-F=0.73249

set.seed(3)
adonis(his.microb.edit ~ bmi.class, permutations=1000) 
  #PERMANOVA=0.3357, pseudo-F=1.0772

set.seed(3)
adonis(his.microb.edit ~ bmi.class2, permutations=1000) 
  #PERMANOVA=0.3017, pseudo-F=1.1061

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 2.747097
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 56
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6819666
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:31), 4])
table(test3[c(32:63), 4])
#Group1 (Higher than median), obese = 18 and non-obese = 13
#Group2 (Lower than median), obese = 20 and non-obese = 12

group1 <- c(18, 13)
group2 <- c(20, 12)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group1", "group2")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.08
## CI = 0.72, 1.61
## p-value = 0.719

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 1.543046
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:31), 2]) #higher group
table(test4[c(32:63), 2]) #lower group
#Group1 (Higher than median), obese = 22 and non-obese = 9
#Group2 (Lower than median), obese = 16 and non-obese = 16

group1 <- c(22, 9)
group2 <- c(16, 16)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.70
## CI = 0.47, 1.07
## p-value = 0.089

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

library(AUCRF)

#Create Obese.num group
edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
obese <- factor(edit.metadata2$obese.num)

#generate test set
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), his.microb.edit)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 3 Measures, AUCopt = 0.7526
plot(fit)

```
<br><br>

##### **Goodrich et al. 2014:**  
```{r, eval=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/twinsUK")

#Read in and match metadata to microbiome data
metadata <- read.csv("TwinsUKStudy_metadata.csv")
rownames(metadata) <- metadata[, 1]
shared.data <- read.table("combined.tx.1.subsample.shared", header=T)
rownames(shared.data) <- shared.data[, 2]
microbiome <- shared.data[, -c(1:3)]
keep  <- rownames(metadata)

microbiome<- microbiome[keep, ]

rm(good.metadata, shared.data, keep)

#generate alpha diversity measures with vegan
library(vegan)
H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
alpha.diversity <- as.data.frame(cbind(H, S, J))
alpha.test <- alpha.diversity

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.table("phyla.txt", header=T)
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(microbiome)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- microbiome
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Chloroflexi", "Deferribacteres", "Elusimicrobia", "Fusobacteria", "Gemmatimonadetes", "Lentisphaerae", "OD1", "Planctomycetes", "Spirochaetes", "SR1", "Synergistetes", "Tenericutes")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4:6, 8:12, 14:17)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100

#Create BMI groups
metadata$BMI.class[metadata$body_mass_index_s<=24] <- "Normal"
metadata$BMI.class[metadata$body_mass_index_s>24 & metadata$body_mass_index_s<30] <- "Overweight"
metadata$BMI.class[metadata$body_mass_index_s>=30 & metadata$body_mass_index_s<40] <- "Obese"
metadata$BMI.class[metadata$body_mass_index_s>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"

#Create a column with obese and extreme obese as single entity
metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"

#Get paitent demographics to be tested
bmi <- metadata$body_mass_index_s
obese <- factor(metadata$obese)
bmi.class <- factor(metadata$BMI.class)
bmi.class2 <- factor(metadata$BMIclass2)

###########################################################################
######### First Level Analysis & Alpha Diversity with BMI #################
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.6671
anova(lm(H ~ bmi.class)) #P-value=0.9732
anova(lm(H ~ bmi.class2)) #P-value=0.8987

anova(lm(S ~ obese)) #P-value=0.9073
anova(lm(S ~ bmi.class)) #P-value=0.9124
anova(lm(S ~ bmi.class2)) #P-value=0.8202

anova(lm(J ~ obese)) #P-value=0.6319
anova(lm(J ~ bmi.class)) #P-value=0.943
anova(lm(J ~ bmi.class2)) #P-value=0.8394

##linear correlations
summary(lm(H ~ bmi)) #P-value=0.8662, R2=5.63x10-5
summary(lm(S ~ bmi)) #P-value=0.9142, R2=2.302x10-5
summary(lm(J ~ bmi)) #P-value=0.8582, R2=6.332x10-5

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

anova(lm(bacter ~ obese)) #P-value=0.6661
anova(lm(bacter ~ bmi.class)) #P-value=0.3184
anova(lm(bacter ~ bmi.class2)) #P-value=0.3184

anova(lm(firm ~ obese)) #P-value=0.9688
anova(lm(firm ~ bmi.class)) #P-value=0.399
anova(lm(firm ~ bmi.class2)) #P-value=0.2455

anova(lm(BFratio ~ obese)) #P-value=0.8709
anova(lm(BFratio ~ bmi.class)) #P-value=0.7915
anova(lm(BFratio ~ bmi.class2)) #P-value=0.5976

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
adonis(microbiome ~ obese, permutations=1000) 
#PERMANOVA=0.004995, pseudo-F=2.9308

set.seed(3)
adonis(microbiome ~ bmi.class, permutations=1000) 
#PERMANOVA=0.005994, pseudo-F=1.8853

set.seed(3)
adonis(microbiome ~ bmi.class2, permutations=1000) 
#PERMANOVA=0.007992, pseudo-F=2.1453

###########################################################################
############ Relative Risk#################################################
###########################################################################

library(epiR)

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity
median(H) # 2.725646
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

#OTU Richness
median(S) # 73
alpha.test <- within(alpha.test, {S.cat = ifelse(S <= median(S), "less", "higher")})

#Evenness
median(J) # 0.6311328
alpha.test <- within(alpha.test, {J.cat = ifelse(J <= median(J), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
S.cat <- alpha.test$S.cat
J.cat <- alpha.test$J.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, S.cat, J.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
table(test3[c(1:253), 4])
table(test3[c(254:507), 4])
#Group1 (Higher than median), obese = 47 and non-obese = 206
#Group2 (Lower than median), obese = 56 and non-obese = 198

group1 <- c(47, 206)
group2 <- c(56, 198)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 1.19
## CI = 0.84, 1.68
## p-value = 0.332

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

median(BFRatio$BFRatio) # 0.5304444
BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
table(test4[c(1:253), 2]) #higher group
table(test4[c(254:507), 2]) #lower group
#Group1 (Higher than median), obese = 52 and non-obese = 201
#Group2 (Lower than median), obese = 51 and non-obese = 203

group1 <- c(52, 201)
group2 <- c(51, 203)
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

epi.2by2(r.test, method="cohort.count")
## Risk Ratio = 0.98
## CI = 0.69, 1.38
## p-value = 0.894

###########################################################################
############ Classification using AUCRF####################################
###########################################################################

library(AUCRF)

#Create Obese.num group
metadata$obese.num[metadata$obese=="No"] <- 0
metadata$obese.num[metadata$obese=="Yes"] <- 1
obese <- factor(metadata$obese.num)

#generate test set
# get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) # list of 8 Measures, AUCopt = 0.6767159
plot(fit)

#Try AUCRF with only phyla and alpha diversity measures in the testset
testset <- cbind(obese, H, S, J, phyla.table.rel.abund)
set.seed(3)
fit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
summary(fit) 
# list of 7 Measures, AUCopt = 0.538138
# Proteobacteria, unclassified, Verrucomicrobia
plot(fit)
```
<br><br>

##### **Escobar et al. 2014:**  
```{r, eval=FALSE}

###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/columbian/")

columbian.microb <- read.table("columbian.0.03.subsample.shared", header=T)
metadata <- read.csv("columbian_dataset.csv")

#Organize the microbiome shared data
rownames(columbian.microb) <- columbian.microb[, 2]
columbian.microb <- columbian.microb[, -c(1:3)]

#Organize the metadata
rownames(metadata) <- metadata[, 5]

#Get only data that we are interested in
edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]

#Sort metadata into the same order as the microbiome data
order1 <- rownames(columbian.microb)
edit.metadata2 <- edit.metadata[order1, ]

#Get alpha diversity of the samples
library(vegan)

H <- diversity(columbian.microb)
S <- specnumber(columbian.microb)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(columbian.microb)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- columbian.microb
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(4:5, 7:9)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Create Obese Yes/No groups
edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"

#Create BMI groups
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s<=24] <- "Normal"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=30 & edit.metadata2$body_mass_index_s<40] <- "Obese"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=40] <- "Extreme Obesity"

#Create BMI groups2
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s<=24] <- "Normal"
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>=30] <- "Obese"

#Get paitent demographics data to be tested 
bmi <- edit.metadata2$body_mass_index_s
obese <- factor(edit.metadata2$obese)
bmi.class <- factor(edit.metadata2$BMI.class)
bmi.class2 <- factor(edit.metadata2$BMI.class2)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

anova(lm(H ~ obese)) #P-value=0.8636
anova(lm(H ~ bmi.class)) #P-value=0.9584
anova(lm(H ~ bmi.class2)) #P-value=0.9584

anova(lm(S ~ obese)) #P-value=0.1947
anova(lm(S ~ bmi.class)) #P-value=0.1432
anova(lm(S ~ bmi.class2)) #P-value=0.1432

>>>>

---
title: "Bacterial Microbiome and Obesity Meta-analysis Supplemental"
author: "Marc Sze"
date: "January 26, 2016"
output: 
  html_document:
    keep_md: yes
bibliography: metaAnalysis.obesity.microbiome.bib
---
<br><br>

```{r data_table_loading, include=FALSE}
setwd("C:/Users/marc/Desktop/obesity2/HMP.analysis")
